---
title: "!~ Exemplary Donation Analysis (EDA) ~!"
author: "Sam Caldwell"
date: "23 May 2018"
output:
  html_document:
    number_sections: true
    toc: true
    code_folding: hide
    theme: cosmo
    highlight: tango
    fig_width: 7
    fig_height: 4.5
---

## Summary
Being a bit of a late comer to the DonorChose.org project, I thought I'd try to do my best to add a fun, new flavour to the analysis already submitted - hopefully by adding some new metrics, data manipulations and plot types to the mix. For such an open-ended task with so many interesting things to look at, I'd thought I'd focus on looking at a few different analysis questions like:

<ul>
<li> How many donors make repeat donations?</li>
<li> What value are donors that make repeat donations?</li>
<li> At what sort of time intervals are repeated donations made?</li>
<li> Are there trends in these intervals?</li>
</ul>

I think these questions could fill hundreds of kernels, but let's give it a crack for now!

## Setup

```{r,message=FALSE,warning=FALSE}
library(readr)
library(tidyverse)
library(leaflet)
library(highcharter)
library(lubridate)

df_donors <- read_csv("data/Donors.csv")
df_donations <- read_csv("data/Donations.csv")
#df_projects <- read_csv("data/Projects.csv")
#df_resources <- read_csv("data/Resources.csv")
df_schools <- read_csv("data/Schools.csv")
#df_teachers <- read_csv("data/Teachers.csv")

data("usgeojson")
data("uscountygeojson")
df_mapdata_states <- get_data_from_map(usgeojson)
df_mapdata_counties <- get_data_from_map(uscountygeojson)

df_state_mapping = df_mapdata_states %>%
  transmute(state_code = postalcode, state_name = name) %>%
  distinct(state_code, state_name)

df_county_mapping = df_mapdata_counties %>%
  transmute(county_code = `code`, county_name = name) %>%
  distinct(county_code, county_name)

```

## Donators'

DonorsChose.org's reach is clearly widespread and impactful across the country, donations are made to large numbers of schools all over the country, as can be seen by the county map below.

```{r}
df_schools_joined <- df_schools %>%
  left_join(df_state_mapping, by = c('School State' = 'state_name')) %>%
  left_join(df_county_mapping, by = c('School County' = 'county_name'))
df_schools_num_county = df_schools_joined %>%
  group_by(county_code) %>%
  summarise(num_schools = n()) %>%
  ungroup() %>%
  drop_na() %>%
  right_join(df_county_mapping) %>%
  replace_na(list(num_schools = 0))

highchart() %>%
  hc_add_series_map(uscountygeojson,  df_schools_num_county, value = "num_schools",
           joinBy = c("code", "county_code"), name = "Number of Schools",
           borderColor = "transparent") %>%
  hc_colorAxis(dataClasses = color_classes(c(1, 10, 100, 1000, max(df_schools_num_county$num_schools)))) %>%
  hc_title(text = "Where the schools at!?") %>%
  hc_legend(layout = "vertical", align = "right",
            floating = TRUE, valueDecimals = 0)
```

But how and when these donations are being made is the key interest point. As noted below, I'll first focus on the donors. Firstly we create a master table using the donations and donors tables to produce metrics at the donor level of analysis to allow slicing and dicing further on.

# Setup master table

```{r}
df_donations_summary = df_donations %>%
  group_by(`Donor ID`) %>%
  arrange(`Donation Received Date`) %>%
  mutate(time_between_donations = round(
    as.numeric(`Donation Received Date`-lag(`Donation Received Date`), units = 'days'))) %>%
  summarise(num_projects = n_distinct(`Project ID`), 
            num_donations = n(), 
            sum_donations = sum(`Donation Amount`),
            first_donation = min(`Donation Received Date`),
            last_donation = max(`Donation Received Date`),
            mean_donation_gap = mean(time_between_donations, na.rm = TRUE),
            max_donation_gap = max(time_between_donations, na.rm = TRUE)) %>%
  mutate(repeat_donations = if_else(num_donations > 1, 1, 0), 
         repeat_projects = if_else(num_projects > 1, 1, 0),
         average_donation = sum_donations/num_donations) %>%
  left_join(df_donors, by = "Donor ID")
```

# Distribution of donors

The brief suggests the organisation is interested in marketing new projects to existing donors, as a result we'll start by compiling the distribution of donors against the number of unique projects they donated to.

```{r}
df_donations_dist = df_donations_summary %>%
  group_by(num_projects) %>%
  summarise(num_donors = n(), sum_donations = sum(sum_donations)) %>%
  mutate(average_donation = sum_donations/num_donors)

df_donations_dist %>%
  hchart('column', hcaes(x = num_projects, y = num_donors)) %>%
  hc_navigator(enabled = TRUE, xAxis = list(labels = list(enabled = FALSE)))
```

Unfortunately the massive peak at 1 project and the long tail dominates the plot - although this does show us the vast majority of donors only donate to one project and that we have some golden donors contributing to huge numbers of projects. Moving slider along the bottom lets us focus on other specific ranges, however for convenience let's focus on an area of interest.

```{r}
df_donations_dist %>%
  filter(num_projects< 10) %>%
  hchart('column', hcaes(x = num_projects, y = num_donors)) %>%
  hc_navigator(enabled = TRUE, xAxis = list(labels = list(enabled = FALSE)))
```
Here we can see the number of donors falling away pretty exponentially compared to the number of projects, overall around 600k donors out of 2 million donors contribute to more than one project - with over half of this amount either 2 or 3 projects.

We could have a hunch that our marketing team want to hunt our multiple project donators for a reason though - let's look at how much on average these groups donate overall!

```{r}
df_donations_dist %>%
  filter(num_projects < 100) %>%
  hchart('column', hcaes(x = num_projects, y = average_donation)) %>%
  hc_navigator(enabled = TRUE, xAxis = list(labels = list(enabled = FALSE)))
```

Boomfa - look at that climbing staircase! - the more different projects people donate to, the more they will donate overall. The average seems to increase steadily (around `$100` per project - perhaps the minimum donation amount?) and then plateaus areound `$8k` per donors.

# Multiple donators map

Clearly those donating to multiple projects are of interest, here is a plot of the states we could find them.
```{r}
df_state_summary = df_donations_summary %>%
  group_by(`Donor State`) %>%
  summarise(prop_rep_donations = 100*sum(repeat_donations)/n(),
            prop_rep_projects = 100*sum(repeat_projects)/n()) %>%
  inner_join(df_state_mapping, by = c('Donor State' = 'state_name')) %>%
  drop_na()
    
highchart() %>%
  hc_add_series_map(usgeojson, df_state_summary, value = "prop_rep_projects",
                    joinBy = c("postalcode", "state_code"), name = "Proportion Donating Multiple Projects",
                    borderColor = "transparent") %>%
  hc_tooltip(valueDecimals = 1, valueSuffix = "%") %>%
  # hc_colorAxis(dataClasses = color_classes(
  #   seq(min(df_state_summary$prop_rep_projects), 
  #       from = max(df_state_summary$prop_rep_projects), length.out = 6))) %>%
  hc_colorAxis(stops = color_stops(6)) %>%
  hc_legend(layout = "vertical", align = "right",
            floating = TRUE, valueDecimals = 0, valueSuffix = "%") %>%
  hc_title(text = "Proportion of Donors Donating to Multiple Projects by State")
```
There is a pretty even spread across the states - most around 23%, with California the highest at 27% and North Dakota the lowest at 19.3%

# Multiple donators over time
Next we look at when the donors first made a donation and compare them in terms of whether they supported another project. Grouping these donors by month, how has the proportion of donors supporting multiple projects compare and how does it compare to the proportion of repeated donations (not necessarily to a new project)?

```{r}
df_monthly_summary = df_donations_summary %>%
  mutate(first_month = as.Date(floor_date(first_donation, "month"))) %>%
  #mutate(first_month =  format(as.Date(first_donation), "%Y-%m")) %>%
  group_by(first_month) %>%
  summarise(prop_rep_donations = 100*sum(repeat_donations)/n(),
            prop_rep_projects = 100*sum(repeat_projects)/n()) %>%
  ungroup() %>%
  gather(variable, value, -first_month)

df_monthly_summary %>%
  hchart("line", hcaes(x = first_month, y = value, group = variable)) %>%
  hc_navigator(enabled = TRUE) %>%
  hc_rangeSelector(enabled = TRUE) %>%
  hc_tooltip(shared = TRUE, valueDecimals = 1, valueSuffix = "%") %>%
  hc_yAxis(title = list(text = "Proportion of Donors %")) %>%
  hc_xAxis(title = list(text = "Month of First Donation")) %>%
  hc_title(text = "Proportion of Donors Making Repeat Donations by Month of First Donation")
```
Both proportions follow eachother closely, although that gap seems to be widening of late - suggesting that if people are making repeat donations then more are to the same project. 

There is a clear downwards trend for both, which could be explained by the donors having less chance to repeat a donation or that our the first donors were the most generous in terms of supporting multiple projects.

# States over time
Did someone say three dimensions!? Let's use a heatmap to combine both findings - is our monthly trend the same across all states?
```{r}
df_monthly_state_summary = df_donations_summary %>%
  mutate(first_month = as.Date(floor_date(first_donation, "month"))) %>%
  group_by(first_month, `Donor State`) %>%
  summarise(prop_rep_donations = 100*sum(repeat_donations)/n(),
            prop_rep_projects = 100*sum(repeat_projects)/n()) %>%
  ungroup()

df_monthly_state_summary %>%
  hchart("heatmap", hcaes(y = first_month, x = `Donor State`, value = prop_rep_projects)) %>%
  hc_colorAxis(stops = color_stops(10)) %>%
  hc_tooltip(shared = TRUE, valueDecimals = 1, valueSuffix = "%") %>%
  hc_yAxis(showLastLabel = FALSE, showFirstLabel = FALSE) %>%
  hc_size(height = 800)
```
Although slightly hard to read, if we follow the states upwards through time we see that most show a similar decreasing trend of people making repeat donations. But is this simply because newer donors have had less of a chance to make another donation? Let's try to find out!

## Donation times
If we can get an idea of when donors make their next donations, we can look at which donors are more likely to donate again soon. Firstly let's look at the distribution of times between successive donations by donors.

# Distribution of mean
Our donation summary dataset from before conveniently took the difference in times between successive donations and aggregated them by donor. Below is the distribution of those donors by that average time difference.
```{r}
df_donations_summary %>%
  mutate(mean_donation_gap = round(mean_donation_gap)) %>%
  group_by(mean_donation_gap) %>%
  summarise(num_donors = n()) %>%
  drop_na() %>%
  hchart("line", hcaes(x = mean_donation_gap, y = num_donors)) %>%
  hc_navigator(enabled = TRUE, xAxis = list(labels = list(enabled = FALSE))) %>%
  hc_yAxis(title = list(text = "Number of Donors")) %>%
  hc_xAxis(title = list(text = "Mean Time Between Donations (days)"))
```
This shows that the vast majority of the donors make repeated donations within the same day, although this is just 80k of the 550k donors. 

Zooming in with the navigator, or filtering the plot as below, we can see small peaks at around 181/364 days - these are probably donors on some kind of yearly/half yearly donation schedules.

```{r}
df_donations_summary %>%
  mutate(mean_donation_gap = round(mean_donation_gap)) %>%
  group_by(mean_donation_gap) %>%
  summarise(num_donors = n()) %>%
  drop_na() %>%
  filter(mean_donation_gap > 1, mean_donation_gap<500) %>%
  hchart("line", hcaes(x = mean_donation_gap, y = num_donors)) %>%
  hc_navigator(enabled = TRUE, xAxis = list(labels = list(enabled = FALSE))) %>%
  hc_yAxis(title = list(text = "Number of Donors")) %>%
  hc_xAxis(title = list(text = "Mean Time Between Donations (days)"))
```


The maximum time between shows a slightly more spread distribution. From this we might be able to deduce a sort of probability of a repeat donation for a given donor, based on how long it's been since their last donation.
```{r}
df_donations_summary %>%
  mutate(max_donation_gap = round(max_donation_gap)) %>%
  group_by(max_donation_gap) %>%
  summarise(num_donors = n()) %>%
  drop_na() %>%
  filter(is.finite(max_donation_gap)) %>%
  hchart("line", hcaes(x = max_donation_gap, y = num_donors)) %>%
  hc_navigator(enabled = TRUE, xAxis = list(labels = list(enabled = FALSE))) %>%
  hc_yAxis(title = list(text = "Number of Donors")) %>%
  hc_xAxis(title = list(text = "Max Time Between Donations (days)"))
```

# Map
Now let's look at how this mean metric varies over the states. 
```{r}
df_state_time = df_donations_summary %>%
  group_by(`Donor State`) %>%
  summarise(mean_donation_gap = mean(mean_donation_gap, na.rm = TRUE)) %>%
  inner_join(df_state_mapping, by = c('Donor State' = 'state_name')) %>%
  drop_na()

highchart() %>%
hc_add_series_map(usgeojson, df_state_time, value = "mean_donation_gap",
      joinBy = c("postalcode", "state_code"), name = "Mean time between donations",
      borderColor = "transparent") %>%
hc_tooltip(valueDecimals = 1) %>%
hc_colorAxis(stops = color_stops(6)) %>%
hc_legend(layout = "vertical", align = "right",
          floating = TRUE, valueDecimals = 0)  
```


## Wrapping up
I'm sure this hasn't scratched the surface of what can be discovered, but I hope it has sparked some interest for others. Happy wrangling!


